---
title: "Week9"
output:
  html_document:
    df_print: paged
editor_options: 
  chunk_output_type: console
---

In this session, we will implement the fitting procedure for a generalized linear model and compare our results with those produced by R’s glm function. In practice, you would typically rely on R's `glm` function to fit a generalized linear model, but building a simple implementation from scratch can help deepen your understanding of how the method works.  

## Logistic regression

We will use a simulated dataset of genetic study to demonstrate logistic regression. We first generate genotype $G \sim Bin(n, f)$ by Hardy-Weinberg principle for $n = 500$ people and the allele frequency $f = 0.3$.

```{r}
set.seed(1)
n = 500
maf = 0.30
G = rbinom(n, 2, maf)
```

Next, for each individual we generate the binary disease status $Y_i$ based on the genotype $G_i$ with a logistic link, where
$Y_i \sim Ber(p_i)$ and $log(\frac{p_i}{1-p_i}) = \beta_0 + G_i \beta_1$. Note that $\frac{p_i}{1-p_i}$ is the odds ratio (probability of having disease vs probabily of not having disease), and we link this linearly to the genotype. We introduce $\eta_i = \beta_0 + G_i \beta_1$, then $p_i = \frac{exp(\eta_i)}{1+exp(\eta_i)}$. We use vectorization to efficiently simulate the calculate all terms.

```{r}
beta0 = -2        # baseline log-odds
beta1_true = 0.5  # log-odds ratio per allele
eta = beta0 + beta1_true * G
p = exp(eta) / (1 + exp(eta))
Y = rbinom(n, 1, p)
```

Fitting a generalized linear model is simple in R. Comparing using `lm` for linear regression, the only additional argument is to specify the distribution to be binomial. R will internally fit the glm, and output the estimate of $\beta_0, \beta_1$, as well as the standard error, test statistics, and p value. 

```{r}
fit_glm = glm(Y~G, family="binomial")
summary(fit_glm)
```

Next, we show how to implement the fitting using Newton's method with Fisher's scoring. Recall from the lecture, the algorithm to fit glm is same for canonical exponential family distributions. We iteratively update $\beta^{new} = \beta^{old} + (X^T W^{old} X)^{-1} X^T (y - \mu^{old})$ where $W$ is a diagonal matrix with each diagonal element $w_i = \frac{\partial \mu_i}{\partial \eta_i}$, and $g(\mu_i) = \eta_i = \beta_0 + G_i \beta_1$.

For logistic regression, in the lecture we derive that $w_i = \mu_i (1-\mu_i)$ and $\mu_i = \frac{exp(\eta_i)}{1+exp(\eta_i)}$. Below is the implementation of the algorithm.

```{r}
my_glm = function(Y, X) {
    n = length(Y)
    
    # initialize beta
    beta = rep(0, ncol(X)) 
    beta_old = beta
    for (i in 1:100) {
        eta = as.vector(X %*% beta) # convert to a vector
        mu = exp(eta) / (1 + exp(eta))
        W = diag(mu * (1 - mu))
        s = t(X) %*% (Y - mu) 
        XtWx = t(X) %*% W %*% X 
        beta = beta_old + solve(XtWx, s)
        # stop criteria
        if (max(abs(beta - beta_old)) < 1e-5) {
            break
        }
        beta_old = beta
    }
    list(beta=beta, mu=mu, XtWx=XtWx)
}

X = model.matrix(~G)
fit = my_glm(Y, X)
fit$beta
```

We are interested in testing whether the genotype is associated with disease status $H_0: \beta_1 = 0$. For glm, there are three types of test: likelihood ratio test, wald test, and score test. They are asymptotically equivalent. We will start with likelihood ratio test.

## Likelihood ratio test

For logistic regression, since $Y_i \sim Ber(p_i)$, the log likelihood of $Y_i$ is $l(Y_i) = Y_i log(p_i) + (1-Y_i) log(1-p_i)$, we then sum over all individuals $\sum_{i=1}^n l(Y_i)$. 

Likelihood ratio test fits two models. First is the reduced model, which only consists of intercept since $\beta_1 = 0$, and the second model is the full model. It then compares the difference of log likelihood of these two model. Under $H_0$, 2 times the difference of the log likelihood (llk_full - llk_reduced) follows chi square distribution with 1 degree of freedom asymptotically. 

We record the value of $\mu$ (same as $p$) when fitting glm to compute the log likelihood. You can also perform LRT using R's anova function.

```{r}
# likelihood ratio test
fit_0 = my_glm(Y, X[, 1, drop=FALSE])
llk_0 = sum(dbinom(Y, prob=fit_0$mu, size=1, log=TRUE))
llk_1 = sum(dbinom(Y, prob=fit$mu, size=1, log=TRUE))
llk_diff = 2*(llk_1 - llk_0)
pchisq(llk_diff, df=1, lower.tail=F)

# using R
fit_0 = glm(Y~1, family="binomial")
fit_1 = glm(Y~G, family="binomial")
anova(fit_0, fit_1, test = "LRT")
```

## Wald test

Wald test only fits the full model, and directly tests whether $\beta_1 = 0$. The test statistics and p value from R's glm function is based on Wald test. 

By mathematical likelihood theory, because $\hat{\beta}$ is the maximal likelihood estimator (MLE), we have $\hat{\beta} -> N(0, I^{-1}(\hat{\beta}))$ under $H_0$, where I^{-1} is the inverse of Fisher information matrix $X^TWX$.  

To test $\beta_2 = 0$, we simply take the second element of $\hat{\beta}$ and the corresponding element (second row, second column) of the inverse of Fisher information matrix $I$, such that $\hat{\beta}_2 -> N(0, I^{-1}(\hat{\beta})_22)$. The test statistics $Z = \frac{\hat{\beta}_2}{I^{-1}(\hat{\beta})_{22}}$ follows standard normal distribution asymptotically.

```{r}
# Wald test
I = fit$XtWx
Z = fit$beta[2] / sqrt(solve(I)[2,2])
2*pnorm(abs(Z), lower.tail=F) 
pchisq(Z^2, df=1, lower.tail=F) 
summary(glm(Y~G, family="binomial"))$coef
```

## Score test

Score test only fits the reduced model to get $\hat{\beta_1}$. It is based on the idea that if $H_0: \beta_2 = 0$ is true (or approximately true), then the estimator $(\hat{\beta_1}, 0)$ should be close to the MLE of fitting the full model. 

Since the score function evaluated at MLE is zero, under $H_0$ evaluating $U(\hat{beta}_1, 0)$ should also be close 0 if $\hat{\beta_2}$ is close to 0. Large deviations from zero indicate evidence to reject the null hypothesis.

By mathematical likelihood theory, we have $U_2 | U_1 = 0 \sim N(0, I_{22} - I_{21} I^{-1}_{11} I_{12})$.

```{r}
# Score test
fit_0 = glm(Y~1, family="binomial")
beta = c(fit_0$coef, 0)
eta = as.vector(X %*% beta)
mu = exp(eta) / (1 + exp(eta))
W = diag(mu * (1 - mu))
s = t(X) %*% (Y - mu)
I = t(X) %*% W %*% X
Z = s[2,1] / sqrt(I[2,2]-I[2,1]*I[1,2]/I[1,1])
S = s[2,1]^2 / (I[2,2]-I[2,1]*I[1,2]/I[1,1])
pchisq(S, df=1, lower.tail=F)

library(statmod)
fit_0 = glm(Y~1, family="binomial")
Z = glm.scoretest(fit_0, G)
pchisq(Z^2, df=1, lower.tail=F)

anova(fit_0, fit_1, test="Rao")
```

## Poisson regression

For Poisson regression, we simulate $Y_i \sim Pois(\mu_i)$ where $log(\mu_i) = \beta_0 + G_i \beta_1$.

```{r}
set.seed(1)
N = 100
G = rnorm(N, 0, 1)
beta0 = -2       
beta1_true = 0.5  
eta = beta0 + beta1_true * G
mu = exp(eta)
Y = rpois(N, mu)
```

The fitting of Poisson regression is similar to fitting the logistic regression. We only need to change the family from "binomial" to "poisson".

```{r}
summary(glm(Y~G, family="poisson"))
```

For implementation of Newton's method, we only need to change how to compute $\mu$ and $W$. For Poisson regression, $\mu = exp(\eta)$ and $w_i = \mu_i$.

```{r}
my_glm = function(Y, X) {
    n = length(Y)
    
    # initialize beta
    beta = rep(0, ncol(X)) 
    beta_old = beta
    for (i in 1:100) {
        eta = as.vector(X %*% beta) # convert to a vector
        mu = exp(eta)  # change for Poisson regression
        W = diag(mu)  # change for Poisson regression
        s = t(X) %*% (Y - mu) 
        XtWx = t(X) %*% W %*% X 
        beta = beta_old + solve(XtWx, s)
        # stop criteria
        if (max(abs(beta - beta_old)) < 1e-5) {
            break
        }
        beta_old = beta
    }
    list(beta=beta, mu=mu, XtWx=XtWx)
}

X = model.matrix(~G)
fit = my_glm(Y, X)
fit$beta
```

The hypothesis testing is also similar. For likelihood ratio test, we change `dbinom` to `dpois`. If using R's anova function, we don't need to change anything.

```{r}
# likelihood ratio test
fit_0 = my_glm(Y, X[, 1, drop=FALSE])
llk_0 = sum(dpois(Y, lambda=fit_0$mu, log=TRUE)) # change for Poisson regression
llk_1 = sum(dpois(Y, lambda=fit$mu, log=TRUE)) # change for Poisson regression
llk_diff = 2*(llk_1 - llk_0)
pchisq(llk_diff, df=1, lower.tail=F)

# using R
fit_0 = glm(Y~1, family="poisson")
fit_1 = glm(Y~G, family="poisson")
anova(fit_0, fit_1, test = "LRT")
```

For Wald test, we don't need to change anything.

```{r}
# Wald test
I = fit$XtWx
Z = fit$beta[2] / sqrt(solve(I)[2,2])
2*pnorm(abs(Z), lower.tail=F) 
pchisq(Z^2, df=1, lower.tail=F) 
summary(glm(Y~G, family="poisson"))$coef
```

For score test, we only need to change $\mu$ and $W$. If using R's function, we don't need to change anything.

```{r}
# Score test
fit_0 = glm(Y~1, family="poisson")
beta = c(fit_0$coef, 0)
eta = as.vector(X %*% beta)
mu = exp(eta) # change for Poisson regression
W = diag(mu) # change for Poisson regression
s = t(X) %*% (Y - mu)
I = t(X) %*% W %*% X
S = s[2,1]^2 / (I[2,2]-I[2,1]*I[1,2]/I[1,1])
pchisq(S, df=1, lower.tail=F)

library(statmod)
fit_0 = glm(Y~1, family="poisson")
Z = glm.scoretest(fit_0, G)
pchisq(Z^2, df=1, lower.tail=F)

anova(fit_0, fit_1, test="Rao")
```
