---
title: "Week5"
output:
  html_document:
    df_print: paged
editor_options: 
  chunk_output_type: console
---

## Population

We first simulate the height of a population of 53,000 individuals from $N(\mu, \sigma^2)$ where $\mu = 170$ and $\sigma = 7$.

```{r}
set.seed(1)
M = 52e3
mu = 170 
sigma = 7
X = rnorm(M, mu, sigma)
```

## Sample

We then randomly draw 100 individuals from the population and calculate the sample mean and the sample variance.

```{r}
set.seed(1)
N = 100
x = sample(X, N)
mean(x)
var(x)
```

We repeatedly draw 100 individuals for 1000 times, each time we  calculate the sample mean $\hat{\mu}$. We plot the histogram of $\hat{\mu}̂$, which is called the sampling distribution. We change the sample size of individuals as $N = 10, 100, 1000$. The red line is the average of $\hat{\mu}$. 

```{r}
set.seed(1)
for (N in c(10, 100, 1000)) {
    res = replicate(1000, {
    x = sample(X, N)
    mean(x)
    })
    hist(res, main=paste0("N = ", N, " mean = ", round(mean(res), 2), " sd = ", round(sd(res),2)))
    abline(v=mean(res), col="red")
}
```

The histogram of $\hat{mu}$ looks like a normal distribution. We standardize $\hat{mu}$ to have mean 0 and variance 1, and use QQplot to compare the quantile of the sampling distribution with the quantile of standard normal distribution $N(0,1)$.

```{r}
set.seed(1)
for (N in c(10, 100, 1000)) {
    res = replicate(1000, {
    x = sample(X, N)
    mean(x)
    })
    qqnorm(scale(res), main=paste0("N = ", N, " mean = ", round(mean(res), 2), " sd = ", round(sd(res),2)))
    abline(0,1)
}
```

## Confidence Interval

Below is the code to illustrate the concept of 95% confidence interval. We repeat the sampling for 100 times. For each draw, we theoretically calculate and plot the 95% confidence interval as $\hat{\mu} - \frac{\hat{\sigma}}{\sqrt{n}} \text{qnorm}(1-\frac{1-0.95}{2}, \hat{\mu} + \frac{\hat{\sigma}}{\sqrt{n}} \text{qnorm}(1-\frac{1-0.95}{2})$. Confidence intervals with blue color indicate that they cover the true value of $\mu = 170$, while confidence intervals with red color indicate that they do not cover the true value of $\mu = 170$. Overall, we see that 6 our of 100 confidence intervals do not cover the true value of $\mu$, which is consistent with the theory.

```{r}
set.seed(10)
N = 100

plot(
  1:100, rep(mu, 100), type = "n",
  ylim = c(165,175),
  ylab = "Mean and 95% Confidence Intervals",
  xlab = "Simulation index",
  main = "Coverage of 95% Confidence Intervals"
)

for (i in 1:100) {
    x = sample(X, N)
    mu = mean(x)
    sigma = sd(x)
    C = sigma / sqrt(N) * qnorm(1-(1-.95)/2)
    lower = mu - C
    upper = mu + C
    col = ifelse(170 >= lower & 170 <= upper, "blue","red")
    segments(i, lower, i, upper, lwd=2, col=col)
    points(i, (lower+upper)/2, pch=16, col=col)
}
legend("topright", legend=c("Contains true mean", "Does not contain"),
  col = c("blue", "red"), lwd=2, pch=16, bty="n")
```

